Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2019-01-27 RNARedPrint.tex
cosmetics

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.html and the instructions for   %%
%% authors page on the biomed central website                      %%https://www.overleaf.com/project/5b98dc8ced90175e9717c15a
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%\documentclass[doublespacing,linenumbers]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass[]{bmcart}
%\documentclass[twocolumn]{bmcart}

@@

%%% Put your definitions there:
\startlocaldefs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Rolf's includegraphicstop
\makeatletter
\newsavebox{\@alignepsbox}
\newlength{\@aligneps}
\newcommand{\includegraphicstop}[2][]{%
\sbox{\@alignepsbox}{\includegraphics[#1]{#2}}%
\setlength{\@aligneps}{-\ht\@alignepsbox}%
\addtolength{\@aligneps}{2ex}%
\raisebox{\@aligneps}{\usebox{\@alignepsbox}}}
\makeatother

@@

\begin{abstractbox}
\begin{abstract}
  \parttitle{Background}
  The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology. Synthetic design approaches profit strongly from effective in-silico methods, \revised{which substantially reduce the need for costly wet-lab experiments.} \\
%
  \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
  design methods: the generation of sequences \revised{that fold well into multiple target structures.} 
  % 
  \revised{Based on constraint networks, our approach \ourprog{}} supports generic
  Boltzmann-weighted sampling, which enables the positive design of RNA sequences
  with specific free energies (for each of multiple, possibly pseudoknotted, target structures) and \GCb-content.
%
  Moreover, we study general properties of our approach empirically and generate
  biologically relevant multi-target Boltzmann-weighted designs for a
  \revised{established} design benchmark. \revised{Our results demonstrate the efficacy and feasibility of the method in practice
  as well as the benefits of Boltzmann sampling over the previously best multi-target sampling strategy---even for the case of negative design of multi-stable RNAs.}
%
  \revised{Besides empirically studies, we finally justify the algorithmic details due to a fundamental theoretic result about multi-stable RNA design, namely the \#P-hardness of the counting of designs.}
%
  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible and---as we demonstrate---effective approach to multi-target RNA design, which promises broad applicability and extensibility.

@@

\begin{figure}
	\includegraphicstop[width=0.8\textwidth]{Figs/Plots/doublestem_weight1-500_mean}
	\caption{\revised{Comparison of free-energy and Boltzmann probability for 10~000 uniform ($\pi=1$; blue dots) and Boltzmann distributed ($\pi=500$; green dots) sequences, targeting a simple structure consisting of two adjacent helices.}}\label{fig:BoltzmannBenefits}
\end{figure}

\revised{First, we show how seed sequences can be generated in a Boltzmann distribution, leading to designs that are substantially more stable that those generated uniformly. As can be seen in Figure~\ref{fig:BoltzmannBenefits} and Supp. Mat. \ref{appsec:immediate-benefits-for-negative-design}, sequences generated in the Boltzmann distribution not only reach lower free-energies than those generated in a uniform setting, but also achieve better Boltzmann probabilities. While the former is expected since the Boltzmann distribution explicitly favors low-energy candidates, the latter is somewhat surprising, since the Boltzmann probability of a target structure could, in principle, decrease under Boltzmann sampling due to the partition function growing faster than the Boltzmann factor. The empirical superiority of Boltzmann sampling appears robust to the target structure length and topology, as demonstrated by prior work~\cite{Reinharz2013}.

\begin{figure}
	  \begin{minipage}{.35\textwidth}\centering 
{\sffamily {\bfseries A} -- 
	Stacking model}\\
	\resizebox{\textwidth}{!}{
		\begin{tabular}{c|c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c}
		$\Delta G$ 	& \Ab\Ub & \Cb\Gb & \Gb\Cb & \Gb\Ub & \Ub\Ab & \Ub\Gb \\[.4em]\hline
			\Ab\Ub &
			-0.18&
			-1.13&
			-1.09&
			-0.38&
			-0.26&
			-0.62
			\\[.4em]
			\Cb\Gb&
			-1.11&
			-2.23&
			-1.89&
			-1.22&
			-1.10&
			-1.44
			\\[.4em]
			\Gb\Ub&
			-0.55&
			-1.26&
			-1.58&
			-0.72&
			-0.50&
			-0.69
			\\
	\end{tabular}}\\[3em]
{\sffamily {\bfseries C} -- Tree widths}\\
\includegraphics[width=\textwidth,trim=350 0 0 50,clip]{Figs/td-widths}
\end{minipage}
	  \begin{minipage}{.6\textwidth}\centering
	{\sffamily {\bfseries B} -- Correlation plot}\\
	  	\includegraphicstop[width=\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor} 
  	\end{minipage}

	\caption{\revised{A fitted energy model based on stacking pairs ({\sffamily\bfseries A}) leads to approximated free-energies that are highly correlated (R=0.99) with free-energies in the Turner energy model ({\sffamily\bfseries B}), yet induces tree widths that are amenable to practical sampling ({\sffamily\bfseries C})}.\label{fig:correlation}}
\end{figure}

However, while in principle feasible, sampling in a Boltzmann distribution directly using the Turner energy model may induce extreme computational demands, with treewidths scaling at least as large as the number of nucleotides in the largest loop. Fortunately, we found that intricacy of the Turner energy model can be circumvented with minimal loss of precision by using a simpler stacking energy model. As shown in Figure~\ref{fig:correlation}, a simple stacking energy model, whose design principles are further described in Supp. Mat.~\ref{appsec:modelparameters}, can be used to approximate the Turner energy model very adequately (correlation coefficient $R=0.99$) in the context of sequence design. Using this simpler model greatly reduces the treewidth, and thus the computational requirements of the whole method even for complex instances.
}

@@

%To reduce the cost of the precomputation, we used a stacking energy model
%as a proxy for the Turner model. Indeed, we observed a strong linear
%dependency between the Turner energy and the
%emphasize that only the simple stacking energies are directly targeted
%(at $\pm$10\% tolerance, except for a few hard instances). Due to our
%linear adjustment, we finally achieve well-controlled energies in the
%realistic Turner energy model.
\begin{figure*}[t]
  \begin{center}
    %{\sf \bfseries A}\
    \includegraphicstop[width=0.97\textwidth]{Figs/energy_shift}\hfill
    %{\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/PKB00211_PKB00239_0_energy_distribution}
  \end{center}
  \caption{%
    Targeting specific energies for pseudoknotted structures using
    multi-dimensional Boltzmann sampling. \textbf{(A)} Linear fits between the
    energies in the stacking model to the realistic pseudoknot energy
    model by Dirks and Pierce (D\&P) for \revised{initially} sampled sequences and both target structures R1 and R2 (shown in B). The good
    match \revised{(respective $R^2$ values of $0.846$ and $0.841$)} enables more efficient targeting of Turner energies based on
    targeting stacking model energies. \textbf{(B)} \revised{Resulting D\&P energy distributions for the two target structures R1 and R2 when aiming for the respective free energies $-30$ and $-20$,
    $-30$ and $-30$, $-25$ and $-25$, $-20$ and $-30$\,kcal/mol. These demonstrate the effectivity of our adaptive
    multi-dimensional Boltzmann sampling procedure, especially by} comparing the distributions to those of uniform and Boltzmann sampled sequences, with homogeneous weights $1$ and
    $e^\beta$, respectively.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
  }
  \label{fig:energydist-pk}
\end{figure*}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.html and the instructions for   %%
%% authors page on the biomed central website                      %%
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\documentclass[doublespacing,linenumbers]{bmcart}% uncomment this for twocolumn layout and comment line below
%\documentclass[]{bmcart}
%\documentclass[twocolumn]{bmcart}

@@

\usepackage{graphicx}
%\usepackage{endfloat}
%\usepackage{caption}
%\captionsetup{labelsep=none,textformat=empty}

%%% Put your definitions there:
\startlocaldefs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Rolf's includegraphicstop
% \makeatletter
% \newsavebox{\@alignepsbox}
% \newlength{\@aligneps}
% \newcommand{\includegraphicstop}[2][]{%
% \sbox{\@alignepsbox}{\includegraphics[#1]{#2}}%
% \setlength{\@aligneps}{-\ht\@alignepsbox}%
% \addtolength{\@aligneps}{2ex}%
% \raisebox{\@aligneps}{\usebox{\@alignepsbox}}}
% \makeatother

@@

\begin{abstractbox}
\begin{abstract}
  \parttitle{Background}
  The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology. Synthetic design approaches profit strongly from effective in-silico methods, \revised{which substantially reduce the need for costly wet-lab experiments.} \\
%
  \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
  design methods: the generation of sequences \revised{that fold well into multiple target structures.} 
  % 
  \revised{Based on constraint networks, our approach \ourprog{}} supports generic
  Boltzmann-weighted sampling, which enables the positive design of RNA sequences
  with specific free energies (for each of multiple, possibly pseudoknotted, target structures) and \GCb-content.
%
  Moreover, we study general properties of our approach empirically and generate
  biologically relevant multi-target Boltzmann-weighted designs for a
  \revised{established} design benchmark. \revised{Our results demonstrate the efficacy and feasibility of the method in practice
  as well as the benefits of Boltzmann sampling over the previously best multi-target sampling strategy---even for the case of negative design of multi-stable RNAs.}
%
  \revised{Besides empirically studies, we finally justify the algorithmic details due to a fundamental theoretic result about multi-stable RNA design, namely the \#P-hardness of the counting of designs.}
%
  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible, and effective approach to multi-target RNA design, which promises broad applicability and extensibility.

@@

\begin{figure}
	\includegraphics[width=0.8\textwidth]{Figs/Plots/doublestem_weight1-500_mean}
	\caption{\revised{Comparison of free-energy and Boltzmann probability for 10~000 uniform ($\pi=1$; blue dots) and Boltzmann distributed ($\pi=500$; green dots) sequences, targeting a simple structure consisting of two adjacent helices.}}
        \label{fig:BoltzmannBenefits}
\end{figure}

\revised{First, we show how seed sequences can be generated in a Boltzmann distribution, leading to designs that are substantially more stable that those generated uniformly. As can be seen in Figure~\ref{fig:BoltzmannBenefits} and Supp. Mat. \ref{appsec:immediate-benefits-for-negative-design}, sequences generated in the Boltzmann distribution not only reach lower free-energies than those generated in a uniform setting, but also achieve better Boltzmann probabilities. While the former is expected since the Boltzmann distribution explicitly favors low-energy candidates, the latter is somewhat surprising, since the Boltzmann probability of a target structure could, in principle, decrease under Boltzmann sampling due to the partition function growing faster than the Boltzmann factor. The empirical superiority of Boltzmann sampling appears robust to the target structure length and topology, as demonstrated by prior work~\cite{Reinharz2013}.}

\begin{figure}
	  \begin{minipage}{.35\textwidth}\centering 
{\sffamily {\bfseries A} -- 
	Stacking model}\\
	\resizebox{\textwidth}{!}{
		\begin{tabular}{c|c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c}
		$\Delta G$ 	& \Ab\Ub & \Cb\Gb & \Gb\Cb & \Gb\Ub & \Ub\Ab & \Ub\Gb \\[.4em]\hline
			\Ab\Ub &
			-0.18&
			-1.13&
			-1.09&
			-0.38&
			-0.26&
			-0.62
			\\[.4em]
			\Cb\Gb&
			-1.11&
			-2.23&
			-1.89&
			-1.22&
			-1.10&
			-1.44
			\\[.4em]
			\Gb\Ub&
			-0.55&
			-1.26&
			-1.58&
			-0.72&
			-0.50&
			-0.69
			\\
	\end{tabular}}\\[3em]
{\sffamily {\bfseries C} -- Tree widths}\\
\includegraphics[width=\textwidth,trim=350 0 0 50,clip]{Figs/td-widths}
\end{minipage}
	  \begin{minipage}{.6\textwidth}\centering
	{\sffamily {\bfseries B} -- Correlation plot}\\
	  	\includegraphics[width=\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor} 
  	\end{minipage}

        \caption{\revised{A fitted energy model based on stacking pairs ({\sffamily\bfseries A}) leads to approximated free-energies that are highly correlated (R=0.99) with free-energies in the Turner energy model ({\sffamily\bfseries B}), yet induces tree widths that are amenable to practical sampling ({\sffamily\bfseries C})}.}
      \label{fig:correlation}
\end{figure}

\revised{%
However, while in principle feasible, sampling in a Boltzmann distribution directly using the Turner energy model may induce extreme computational demands, with treewidths scaling at least as large as the number of nucleotides in the largest loop. Fortunately, we found that intricacy of the Turner energy model can be circumvented with minimal loss of precision by using a simpler stacking energy model. As shown in Figure~\ref{fig:correlation}, a simple stacking energy model, whose design principles are further described in Supp. Mat.~\ref{appsec:modelparameters}, can be used to approximate the Turner energy model very adequately (correlation coefficient $R=0.99$) in the context of sequence design. Using this simpler model greatly reduces the treewidth, and thus the computational requirements of the whole method even for complex instances.
}

@@

%To reduce the cost of the precomputation, we used a stacking energy model
%as a proxy for the Turner model. Indeed, we observed a strong linear
%dependency between the Turner energy and the
%emphasize that only the simple stacking energies are directly targeted
%(at $\pm$10\% tolerance, except for a few hard instances). Due to our
%linear adjustment, we finally achieve well-controlled energies in the
%realistic Turner energy model.
\begin{figure*}[t]
  \begin{center}
    %{\sf \bfseries A}\
    \includegraphics[width=0.97\textwidth]{Figs/energy_shift}\hfill
    %{\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/PKB00211_PKB00239_0_energy_distribution}
  \end{center}
  \caption{%
    Targeting specific energies for pseudoknotted structures using
    multi-dimensional Boltzmann sampling. \textbf{(A)} Linear fits between the
    energies in the stacking model to the realistic pseudoknot energy
    model by Dirks and Pierce (D\&P) for \revised{initially} sampled sequences and both target structures R1 and R2 (shown in B). The good
    match \revised{(respective $R^2$ values of $0.846$ and $0.841$)} enables more efficient targeting of Turner energies based on
    targeting stacking model energies. \textbf{(B)} \revised{Resulting D\&P energy distributions for the two target structures R1 and R2 when aiming for the respective free energies $-30$ and $-20$,
    $-30$ and $-30$, $-25$ and $-25$, $-20$ and $-30$\,kcal/mol. These demonstrate the effectivity of our adaptive
    multi-dimensional Boltzmann sampling procedure, especially by} comparing the distributions to those of uniform and Boltzmann sampled sequences, with homogeneous weights $1$ and
    $e^\beta$, respectively.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
  }
  \label{fig:energydist-pk}
\end{figure*}
